Robustness of MinD oscillation in \textbf{\emph{Escherichia coli}} with diverse cell shapes

### 2014-11-04 paper/paper-one.tex
added my best attempt at putting everything into pnas format, added style file cls file and changes to Makefile Tested everything on crystal.

\documentclass[letterpaper,twocolumn,amsmath,amssymb,pre]{revtex4-1}
\usepackage{graphicx}% Include figure files
\usepackage{color}

@@

\begin{document}
\title{Robustness of MinD oscillation in \emph{Escherichia coli} with
  diverse cell shapes}

\author{Jeff B. Schulte}
\affiliation{Department of Physics, Oregon State University}
\author{Rene W. Zeto}
\affiliation{Department of Physics, Oregon State University}
\author{David Roundy}
\affiliation{Department of Physics, Oregon State University}

\begin{abstract}
  The dynamics of the Min-protein system help \emph{Escherichia coli}
  regulate the process of cell division by identifying the center of
  the cell.  While this system usually exhibits robust polar
  oscillations in a variety of cell shapes, new experiments have shown
  that when the cells are mechanically deformed into wide, flattened
  out, irregular shapes, the spatial regularity of these oscillations
  breaks down. Here we study widely used stochastic and deterministic
  models of Min-system simulation within these new cell shapes.  We
  find that the deterministic model is decidedly inadequate to
  reproduce the experimentally observed behavoir, while the stochastic
  model, based on the same differential equations, is able to do so
  accurately.  We also find that it is the flattening rather than the
  irregularity and asymmetry of the cell shape that causes the
  irregular oscillation behavoir.
\end{abstract}

@@

\section{Introduction}
It is vital that during the process of bacterial cell division a cell
avoid minicelling, or splitting into daughter cells with lopsided
volumes.  Instrumental to this process in \emph{Escherichia coli} is a
long FtsZ polymer chain that develops on the cell wall in the center
region of the cell, helping dictate the plane of
division~\cite{adams2009bacterial, lutkenhaus2007assembly}. Previous
experimental studies have shown that the MinC protein, known to
inhibit the FtZ polymer~\cite{shen2010examination}, exhibits regular
pole to pole oscillatory behavior between both ends of the wild-type
pill-shaped cell.  It thus has a higher time averaged concentration in
the cell poles than in the center region, which aides in preventing
the FtZ from developing in the wrong region.  The MinC is recruited to
these poles by MinD, which itself interacts with another protein,
MinE, in a system which exhibits pole-to-pole oscillatory
behavoir~\cite{hu1999topological, fu2001mine, shapiro2009and,
  yu1999ftsz, raskin1999rapid, meacci2005min, raskin1999minde}.

@@

%% \fixme{There are not any direct comparisons of the deterministic and
%%   stochastic models that I can find other than the fange2006 paper
%%   cited above for the round cells.  Those showed that the
%%   deterministic sims are 'bistable' - one type of initial conditions
%%   (MinD starting on walls on one side) show oscillations, another type
%%   (half cyto is filled with 3/4 of MinD) don't.  Huang in 2004 paper
%%   cited above uses numerical determinisitc and finds that it can show
%%   oscillations in round cells as well, but right now I can't find it
%%   and the journal doesn't give me access to it.  predict oscillations
%%   in perfectly round cells unless noise was added to the simulation,
%%   but could in almost round ellipses.  It seems that once the
%%   stochastic models began to be used, nobody really used the
%%   deterministic model any longer.  Not sure what to say about this in
%%   order to introduce what we do here properly.}

@@

@@

%% Our pill shapes differ from those of Huang \emph{et al.} in that they
%% are cylinders with hemisphere endcaps instead of pure cylindrical
%% shapes.  Our cylinderidrical radius is $0.5\micron$ and the lengths
%% of our cells (measured between the tips of the endcaps) are
%% $5\micron$, $4\micron$, $3\micron$, and $2.5\micron$.

%% \fixme{Is the following Kubitscheck paragraph needed?}
%% Kubitschek has shown in multiple experiments that at the time of cell
%% division cells have a volume that is within a range of roughly
%% $1\micron^3$ to $2\micron^3$~\cite{kubitschek1990cell,
%%   kubitschek1968linear}.  We follow Huang's
%% simulations\cite{huang2003dynamic} and Mannik's experiments and model
%% cells that are slightly larger than this range.

\begin{figure*}
  %\includegraphics[width=\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-exact/plots/image-plot}
  \begin{center}
    \includegraphics[width=0.8\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-exact/plots/single-image-plot}\\
    \vspace{-1.5em}
    \includegraphics[width=0.8\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-full_array/plots/single-image-plot}
    \vspace{-3em}
  \end{center}
  \caption{Images of the concentration of each protein species in a
    natural pill-shaped bacterium at one-second time intervals. The
    upper plots shows results from the deterministic model and the
    lower shows results from the stochastic model.  The order of
    frames is such that individual MinD proteins begin at the bottom
    of the plot (in the MinD:ATP state in the cytoplasm), and progress
    upward until they reach the MinE:MinD:ATP membrane-bound complex.
    At that point, they will spontaneously dissociate into cytoplasmic
    MinE (the top row) and the starting state of cytoplasmic
    MinD:ADP.}.
  \label{image-p}
\end{figure*}

@@

@@

%% is concentrated in the
%% upper portion of the cell and is diffusing downward, where it will
%% react with and stick to wall-bound MinD that is concentrated on the
%% membrane in the lower portion of the cell.  This process is in fact
%% already well on its way, as seen by the significant concentration of
%% wall bound MinD:MinE which is creeping down to the lower corner,
%% removing the MinD from the membrane as it goes.  The removed MinD
%% (cytoplasmic MinD:ADP) diffuses freely for a time before spontenously
%% undergoing nucleotide exchange and shifting back into the MinD:ATP
%% that is able to attach to the membrane.  Diffusing a small distance
%% away from the bottom of the cell will bring the protein into the MinE
%% \fixme{collection ??}, where reactions will prevent it from collecting
%% for any significant time on the membrane.  Diffusing further towards
%% the upper end, however, while the MinE has not yet collected there,
%% will allow MinD:ATP to accumulate with the other MinD:ATP, which is
%% seen at seconds 10 through 15 in Figure~\ref{image-p}.  Eventually the
%% MinE creeps all the way down to the bottom of the cell where it slowly
%% runs out of wall bound MinD to react with, before diffusing back up
%% and starting again on the upper portion of the cell. This is seen in
%% seconds 15 through 20 in the figure.  \fixme{There is another
%%   paragraph after this in the latex that I commented out because I
%%   thought we might be spending too much time on the description}

%%  At 15 seconds, the membrane-bound MinD:ATP has
%% been essentially removed from the top end of the cell, and MinD:ATP
%% has begun to bind to the membrane in the the lower half of the cell.
%% At this stage, there is a high concentration of cytoplasmic MinE at
%% the top of the cell, and by 16 seconds we begin to see the formation
%% of a MinE ring just below the center of the cell.  At 18 seconds, the
%% cell has reached its original state (reversed directionally), with
%% MinD:ATP bound to the membrane on the lower third of the cell, a high
%% concentration of cytoplasmic MinE in the upper half of the cell, and a
%% MinE ring pushing downward on the membrane-bound MinD:ATP.

@@

\section{Comparison of Mannik's Experimental and Our Stadium Shapes}

As described in Section~\ref{sec:model-method-shapes}, we will focus
on simulation of four illustrative flattened cell shapes: two shapes
created to replicate those shapes observed experimentally by
Mannik~\cite{mannik2012robustness} (shape~\emph{A} and
shape~\emph{B}), and for comparison two symmetrical stadium shapes
that have the identical aspect ratio, thickness and volume (stadium
\emph{A} and stadium \emph{B}), shown in Figure~\ref{randst-plot-ave}.
These stadium shapes enable us to distinguish between the effect of
flattening the cell and the cell shape's irregularity and asymmetry.
We note that shape \emph{B} is more irregular than shape \emph{A}, and
in particular features a prominant region of concavity on its
left-hand side.  For each of these shapes (in addition to the
wild-type pill shape discussed previously) we have simulated in excess
of 15000 seconds (over 4 hours) of evolution of the MinD system using
the deterministic version of Huang's model~\cite{huang2003dynamic},
and in excess of 100000 seconds (over 27 hours) using the stochastic
model.

@@

We find that flattened but regular and symmetric cells exhibit MinD
oscillations that are qualitatively similar to the oscillations
observed experimentally (and in our simulation) in irregular and
asymmetric cells.  This demonstrates that asymmetry is not required in
order to induce spatially irregular MinD oscillations.  In addition,
the temporal regularity of oscillations is only moderately affected by
irregularity in the shape of flattened cells.  Thus we conclude
that it is the flattening of the cells rather than their irregular
shape that primarily causes the spatial irregularity of maxima which
is observed by Mannik \emph{et al.}~\cite{mannik2012robustness}.
%% %
%% In order to understand the irregular MinD oscillations observed in
%% experiments with flattened cells, we simulated flattened ``stadium''
%% shapes in order to distinguish between the effect of flattening and
%% the effect of irregular shape.
%% %
%% We found only a moderate difference in the coherence of bipolar
%% oscillations within these stadium shapes and the irregular shapes
%% observed by Mannik \emph{et al.}, and have observed visually the same
%% qualitative spatially irregular formation of maxima in both.  Thus we
%% conclude that the disruption of the MinD oscillations seen by Mannik
%% \emph{et al.}  is largely due to the flattening of the cell shape, as
%% opposed to its irregularity.

\bibliography{paper}

@@

\section*{Appendix}

\section{Below are NOTES for the Writing}

%% mannik B 11500 seconds, mannik A 15600 seconds, stadium A 19120  seconds, pill 19990 seconds
%% = simulation time that deterministic correlation data covers.

>>>>

%% PNAStwoS.tex
%% Sample file to use for PNAS articles prepared in LaTeX
%% For two column PNAS articles
%% Version: Apr 15, 2008

%% BASIC CLASS FILE
\documentclass{pnastwo}

%% ADDITIONAL OPTIONAL STYLE FILES
%\usepackage[dvips]{graphicx}
%\usepackage{pnastwoF}
\usepackage{amssymb,amsfonts,amsmath}

@@

%% OPTIONAL MACRO DEFINITIONS
%\def\s{\sigma}

%%%%%%%%%%%%
%% For PNAS Only:
\url{www.pnas.org/cgi/doi/10.1073/pnas.0709640104}
\copyrightyear{2008}
\issuedate{Issue Date}
\volume{Volume}
\issuenumber{Issue Number}
%\setcounter{page}{2687} %Set page number here if desired
%%%%%%%%%%%%

\begin{document}

\title{Robustness of MinD oscillation in \textbf{\emph{Escherichia coli}} with
  diverse cell shapes}

\author{Jeff B. Schulte\affil{1}{Department of Physics, Oregon State University},
Rene W. Zeto\affil{1}{}
\and
David Roundy\affil{1}{}}

\contributor{Submitted to Proceedings of the National Academy of Sciences
of the United States of America}

@@

\begin{article}
\begin{abstract}
  The dynamics of the Min-protein system help \emph{Escherichia coli}
  regulate the process of cell division by identifying the center of
  the cell.  While this system usually exhibits robust polar
  oscillations in a variety of cell shapes, new experiments have shown
  that when the cells are mechanically deformed into wide, flattened
  out, irregular shapes, the spatial regularity of these oscillations
  breaks down. Here we study widely used stochastic and deterministic
  models of Min-system simulation within these new cell shapes.  We
  find that the deterministic model is decidedly inadequate to
  reproduce the experimentally observed behavoir, while the stochastic
  model, based on the same differential equations, is able to do so
  accurately.  We also find that it is the flattening rather than the
  irregularity and asymmetry of the cell shape that causes the
  irregular oscillation behavoir.
\end{abstract}

\keywords{MinD | oscillation | stochastic | deterministic | simulation}

\abbreviations{SAM, self-assembled monolayer; OTS,
octadecyltrichlorosilane}

\dropcap{I}t is vital that during the process of bacterial cell division a cell
avoid minicelling, or splitting into daughter cells with lopsided
volumes.  Instrumental to this process in \emph{Escherichia coli} is a
long FtsZ polymer chain that develops on the cell wall in the center
region of the cell, helping dictate the plane of
division~\cite{adams2009bacterial, lutkenhaus2007assembly}. Previous
experimental studies have shown that the MinC protein, known to
inhibit the FtZ polymer~\cite{shen2010examination}, exhibits regular
pole to pole oscillatory behavior between both ends of the wild-type
pill-shaped cell.  It thus has a higher time averaged concentration in
the cell poles than in the center region, which aides in preventing
the FtZ from developing in the wrong region.  The MinC is recruited to
these poles by MinD, which itself interacts with another protein,
MinE, in a system which exhibits pole-to-pole oscillatory
behavoir~\cite{hu1999topological, fu2001mine, shapiro2009and,
  yu1999ftsz, raskin1999rapid, meacci2005min, raskin1999minde}.

@@

@@

%% \begin{figure}
%%   \includegraphics[width=\columnwidth]{reactions}
%%   \caption{Reactions included in the model of Huang \emph{et
%%       al.}~\cite{huang2003dynamic}.}\label{fig:reactions}
%% \end{figure}

@@

%% \begin{figure*}
%%   %\includegraphics[width=\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-exact/plots/image-plot}
%%   \begin{center}
%%     \includegraphics[width=0.8\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-exact/plots/single-image-plot}\\
%%     \vspace{-1.5em}
%%     \includegraphics[width=0.8\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-full_array/plots/single-image-plot}
%%     \vspace{-1.5em}
%%   \end{center}
%%   \caption{Images of the concentration of each protein species in a
%%     natural pill-shaped bacterium at one-second time intervals. The
%%     upper plots shows results from the deterministic model and the
%%     lower shows results from the stochastic model.  The order of
%%     frames is such that individual MinD proteins begin at the bottom
%%     of the plot (in the MinD:ATP state in the cytoplasm), and progress
%%     upward until they reach the MinE:MinD:ATP membrane-bound complex.
%%     At that point, they will spontaneously dissociate into cytoplasmic
%%     MinE (the top row) and the starting state of cytoplasmic
%%     MinD:ADP.}.
%%   \label{image-p}
%% \end{figure*}

@@

%% \begin{figure}
%%   \includegraphics[width=\columnwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-full_array/plots/correlation.pdf}
%%   \caption{Temporal correlation function of the total MinD found in
%%     two opposite polar regions of the pill-shaped cell, shown against
%%     the correlation time.  Data for both the the determinisitic and
%%     stochastic models are shown.  The stochastic model shows an
%%     oscillation period of 39.5 seconds and coherence time of 307 seconds.
%%     The correlation functions are scaled to have the same initial
%%     value.}
%%   \label{corr-pill}
%% \end{figure}

@@

@@

%% \begin{figure*}
%%   \centering
%%   \includegraphics[width=0.8\textwidth]{../paper/plot-ave}
%%   \caption{We display here arrows depicting successive maxima in space
%%     and and time overlayed on a color plot of the total MinD density
%%     averaged over the same time period.  The simulation time covered
%%     for the flattened cells is 350 seconds, which is the same
%%     period of time depicted in the experimental data plots of Mannik
%%     \emph{et al.}~\cite{mannik2012robustness}.  For the wild-type pill
%%     shape, we only cover 250 seconds, in order to provide a
%%     useful comparison due to its shorter oscillation period.  The top
%%     row shows plots published by Mannik of the MinD maxima behavoir
%%     and the bottom two rows show our simulations using the stochastic
%%     and deterministic models, respectively.  We simulated
%%     approximations to the two shapes observed by Mannik, which we call
%%     \emph{shape A} and \emph{shape B}.  In addition, we studied two
%%     flattened stadium shapes which we call \emph{stadium A} and
%%     \emph{stadium B} corresponding in aspect ratio and thickness to
%%     the two experimental shapes.  The spatial length scale of all the
%%     figures shown was identical. Each of the flattened cell shapes
%%     uses the same color scale for the number of proteins per unit
%%     area.  Finally, we display the natural pill shape, which was also
%%     featured in Figs.~\ref{image-p} and~\ref{corr-pill}, with a
%%     different color scale to reflect the thicker cell containing more
%%     proteins per cross-sectional area.  }
%%   \label{randst-plot-ave}
%% \end{figure*}

%% \begin{figure}
%%   \includegraphics[width=\columnwidth]{../data/shape-randst/0_25-18_50-18_50-95_00-15_00-full_array/plots/correlation.pdf}
%%   \includegraphics[width=\columnwidth]{../data/shape-stad/0_25-2_35-1_32-0_00-15_00-full_array/plots/correlation.pdf}
%%   \caption{Temporal correlation function of the total MinD found in
%%     two opposite polar regions of the shape \emph{A} (above) and the
%%     stadium \emph{A} (below) cell shapes, shown against the
%%     correlation time.  Data for both the the determinisitic and
%%     stochastic models are shown, and the correlation functions are
%%     scaled to have the same initial value.  For shape \emph{A}, the
%%     stochastic model shows an oscillation period of 53.1 seconds and
%%     coherence time of 246 seconds, so that it takes roughly 4.6 periods
%%     for the behavoir to decohere.  For the stadium \emph{A}, the same
%%     model shows an oscillation period of 51 seconds and coherence time of
%%     323 seconds, so that it takes roughly 6.3 periods for the behavoir
%%     to decohere.}
%%   \label{fig:corr-pancake-A}
%% \end{figure}

\section{Comparison of Mannik's Experimental and Our Stadium Shapes}
As described in Section~\ref{sec:model-method-shapes}, we will focus
on simulation of four illustrative flattened cell shapes: two shapes
created to replicate those shapes observed experimentally by
Mannik~\cite{mannik2012robustness} (shape~\emph{A} and
shape~\emph{B}), and for comparison two symmetrical stadium shapes
that have the identical aspect ratio, thickness and volume (stadium
\emph{A} and stadium \emph{B}), shown in Figure~\ref{randst-plot-ave}.
These stadium shapes enable us to distinguish between the effect of
flattening the cell and the cell shape's irregularity and asymmetry.
We note that shape \emph{B} is more irregular than shape \emph{A}, and
in particular features a prominant region of concavity on its
left-hand side.  For each of these shapes (in addition to the
wild-type pill shape discussed previously) we have simulated in excess
of 15000 seconds (over 4 hours) of evolution of the MinD system using
the deterministic version of Huang's model~\cite{huang2003dynamic},
and in excess of 100000 seconds (over 27 hours) using the stochastic
model.

@@

We find that flattened but regular and symmetric cells exhibit MinD
oscillations that are qualitatively similar to the oscillations
observed experimentally (and in our simulation) in irregular and
asymmetric cells.  This demonstrates that asymmetry is not required in
order to induce spatially irregular MinD oscillations.  In addition,
the temporal regularity of oscillations is only moderately affected by
irregularity in the shape of flattened cells.  Thus we conclude
that it is the flattening of the cells rather than their irregular
shape that primarily causes the spatial irregularity of maxima which
is observed by Mannik \emph{et al.}~\cite{mannik2012robustness}.

\bibliographystyle{plain}
\bibliography{paper-one}
\end{article}

%\section*{Appendix}
\begin{acknowledgments}
This work was partially supported by
Spanish Ministry of Science and Technology Grant BFM2002-02042 (to D.C. and
J.L.R.) and by National Science Foundation Grand DMS-0245242 (to C.F.).
\end{acknowledgments}

\begin{figure*}
  %\includegraphics[width=\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-exact/plots/image-plot}
  \begin{center}
    \includegraphics[width=0.8\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-exact/plots/single-image-plot}\\
    \vspace{-1.5em}
    \includegraphics[width=0.8\textwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-full_array/plots/single-image-plot}
    \vspace{-1.5em}
  \end{center}
  \caption{Images of the concentration of each protein species in a
    natural pill-shaped bacterium at one-second time intervals. The
    upper plots shows results from the deterministic model and the
    lower shows results from the stochastic model.  The order of
    frames is such that individual MinD proteins begin at the bottom
    of the plot (in the MinD:ATP state in the cytoplasm), and progress
    upward until they reach the MinE:MinD:ATP membrane-bound complex.
    At that point, they will spontaneously dissociate into cytoplasmic
    MinE (the top row) and the starting state of cytoplasmic
    MinD:ADP.}.
  \label{image-p}
\end{figure*}

@@
